Individual foraging specialization and success change across experience in a virtual predator-prey system

### 2024-11-26 manuscript/manuscript.Rmd
update manuscript

---
title: "Experience shapes individual foraging specialization and success in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

@@

The capacity of predators to match their tactic to their prey is expected to drive the outcome of predator-prey interactions. To do so, predators need to practice their tactics extensively to develop the skills to be successful hunters, which may result in individual foraging specialization. Yet, there are limited empirical assessments showing links between experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. Here, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual foraging specialization and success in predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience. Prey speed was an important mediator of this relationship, resulting in differences among predators in the development of expertise. Throughout experience, 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by switching between slow and fast speeds. The predators’ strategies were partially matched to the preys’ speed, suggesting that changes in hunting behaviour were driven by repeated encounters with prey. Flexible and specialist foragers achieved similar success overall. Our findings suggest that experience may be mechanism promoting behavioural diversification in predator-prey systems.

@@

Predators express a diverse array of foraging tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing and fast movement speeds are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, and may have larger implications for community and ecosystem dynamics [@wooster.etal2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@dukas2019;@wooster.etal2023]. Through this process, it is hypothesized that individuals will gradually specialize in a foraging tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each attempt [@tinker.etal2009;@dukas2019]. This should occur when predators learn to hunt in environments where resources are predictable (or stable) because they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. Such conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathot.etal2012]. Nonetheless, complex feedback between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@stephens1993;@ishii.shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences are driven by resource fluctuations [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. For example, @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Prey that use fast escapes are also proven to be more difficult to hunt [@walker.etal2005;@kelley.magurran2011;@martin.etal2022], and predators that exclusively encounter such prey should benefit from learning to specialize in fast pursuits. If predators that hunt in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both attain similar hunting success and coexist [@phillips.etal2017]. Therefore, empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Similar to agent-based simulations, online videogames can be useful systems to mitigate these challenges and test specific ecological hypotheses. They provide controlled virtual environments with the advantage of having real players that interact in the virtual space. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@fraserfranco.etal2022, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. In a different system, @lymbery.etal2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how repeated experience shapes predator foraging specialization and success using data from players in *Dead by Daylight*. Following @fraserfranco.etal2022 we use the predators' mean movement speed along a slow-fast continuum as a proxy for the foraging tactic (ambush-cursorial), and define the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual foraging specialization (e.g. always using a cursorial tactic). First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that the predators' success will increase with experience up to a certain level where it will stagnate. We expect this pattern to vary among individuals due to differences in prey encounters. Second, we assess how foraging specialization emerges with experience. If all predators encounter similar groups of prey, we predict that individuals (and thus the population) should all specialize in similar speeds. In contrast, if all predators encounter varying groups of prey, then they should all converge towards flexible speeds. In both scenarios, differences among individuals in IIV across experience should be low (i.e. similar individual foraging specialization), whereas the population variance would either decrease (specialization) or increase (flexibility). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, we expect predators that encountered similar groups of prey across experience to specialize in similar speeds, while predators that encountered heterogeneous groups across matches should adopt flexible movement speeds, resulting in an increase in among individual differences in IIV with experience. If we detect such prey-dependent fine-tuning with experience, then all hunters along the flexible-specialist continuum should attain equal success.

@@

*DBD* is a survival asymmetric multiplayer online game (i.e. a game where the gameplay mechanics differ between two groups) developed by Behaviour Interactive Inc, in which players can play either as a predator or a prey. Every match includes only one predator and four prey. The objective of the predator is to hunt and capture the prey, and the objective of the prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group in a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that have not escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can choose an avatar with abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments are composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling them to have visual cues of the generators [@fraserfranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during the study period.

@@

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We only analyzed matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when mean distances traveled per second (i.e. speed) were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as intended. We then sampled players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience along with its speed, the mean speed of the groups of prey the player encountered, and their mean rank (a proxy for prey skill). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank), and failing to account for it would prevent us from detecting a change in the predator's foraging success with experience.

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. We did not account for matches where predators played as the prey. We also categorized predators for each match based on their cumulated experience. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 499). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

We recognize that we could have biased our analyses by retaining only individuals who played for at least 300 matches. For example, these individuals might be experienced videogame players and could thus already be playing like experts in their first matches in *DBD*. To verify that our sample was not biased, we compared a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our sampled population. We then took the first 20 matches played by these players, including those from our sampled population, and compared their mean speed using a Bayesian hierarchical linear model. We found that all four groups had similar mean speeds as predators (Appendix 2: Table S1-S2), which indicates an absence of bias due to data sampling.

@@

### Software and computer specifications

All models were fitted under a Bayesian framework in R (version 4.1.2) using Markov chain Monte Carlo (MCMC) sampling with the package "brms" version 2.16.3 [@burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.4.0 [@gabry.cesnovar2021] as the back-end for estimation (cmdstan installation version 2.28.2). The models were run on Cedar (Operating system: CentOS Linux 7), a computer cluster maintained by the Digital Research Alliance of Canada (https://docs.alliancecan.ca/wiki/Cedar). Each required 64GB of RAM with 48 cores to compile within 5 days.

### Development of expertise

We tested how prey influenced the development of predator expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were reproductions of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

We computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Hunting success was estimated as the probability of consuming the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link. We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$), a positive Gaussian prior ($N(1,\ 0.5^{2}))$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

@@

We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

After inspection of the results, we found that key parameter values for our hypotheses were in the direction of our predictions, but they were largely driven by the very slow speed of one predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix S3.

### Markov Chain Monte Carlo settings

We parametrized the GAMMs and the MDHGLM to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table 1), and 1500 iterations with a thinning set to four for the other additive models. We set the first 500 iterations in each model as warm ups. We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 iterations used as warm ups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

@@

## Development of expertise

@@

# Open percentages table
path <- file.path(here::here("outputs", "03_outputs_model-processing"))
percent_tab1 <- data.table(readRDS(file.path(path, "GAMM_percentages.rds")))
percent_tab2 <- data.table(readRDS(file.path(path, "DHMLM_percentages.rds")))

#fit1 <- readRDS(here::here("outputs", "01_outputs_models", "A2_GAMM-rank.rds"))
#fit2 <- readRDS(here::here("outputs", "01_outputs_models", "A2_GAMM-speed-rank.rds"))
#
#fit1
#fit2
```

Out of all five GAMM models, two predicted the data the best and achieved similar expected log pointwise densities (elpd). Both models accounted for the prey group's rank and speed (Table 1). Models in which these effects were not accounted for resulted in no change in hunting success with experience (i.e. expertise) for the average individual (results not shown). Accounting for the prey rank resulted in a concave-shaped relationship, with the highest success ranging between 100 and 400 matches (Figure 1A). In the model where we additionally included the prey's speed, the effect of experience on hunting success for the average individual followed a diminishing returns curve, with predators optimizing their success after playing ~300 matches (Figure 1C). The curve shows there was a 38% increase in the probability of consuming all prey for the average individual between the first and the ~330$^{th}$ match where success reached a plateau (Figure 1C).

```{r table1, ft.align = "left"}

# ===========================================================================
# 1. Import the model fit
# ===========================================================================

# Open the loo table
path <- file.path(here::here("outputs", "02_outputs_model-validation"))
loo_tab <- data.table(readRDS(file.path(path, "GAMM_loo-cv-table.rds")))

# Add model column
loo_tab[
  , model := c(
    "predator xp + ID smoothers + prey rank + prey speed",
    "ID smoothers + prey rank + prey speed",
    "ID smoothers + prey rank",
    "predator xp + ID smoothers + prey rank",
    "predator xp + prey rank"
    )
 ]

# Reorder columns
loo_tab <- loo_tab[, c(9, 1:8)]

# ===========================================================================
# ===========================================================================

# ===========================================================================
# 2. Compute the table using flextable
# ===========================================================================

# Prepare the table parameters ----------------------------------------------

# Custom header
my_header <- data.frame(
   col_keys = c("model",
                "elpd_diff",
                "se_diff",
                "elpd_loo",
                "se_elpd_loo"),
   line1 = c("model",
             "elpd \ndifference",
             "sd \ndifference",
             "elpd loo \nvalue",
             "elpd loo \nstandard error"),
   stringsAsFactors = FALSE
)

# Custom theme
my_theme <- function(x, ...) {
   x <- colformat_double(x, big.mark = " ",
                         decimal.mark = ".",
                         digits = 2)
   x <- set_table_properties(x, layout = "autofit")
   x <- border_remove(x)
   std_border <- fp_border(width = 1, color = "black")
   x <- hline_top(x, part = "all", border = std_border)
   x <- hline_bottom(x, part = "all", border = std_border)
}

# Create the table ----------------------------------------------------------

tab1 <- loo_tab %>%
  select(model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  flextable(col_keys = my_header$col_keys) %>%
  set_header_df(mapping = my_header, key = "col_keys") %>%
  my_theme() %>%
  align(align = "left", part = "all", j = 1) %>%
  align(align = "center", part = "all", j = 2) %>%
  align(align = "center", part = "all", j = 3) %>%
  align(align = "center", part = "all", j = 4) %>%
  align(align = "center", part = "all", j = 5) %>%
  footnote(
    i = 1, j = 1,
    part = "header",
    value = as_paragraph(
      as_sup("a "),
      "'elpd' refers to the expected log pointwise density and is the value chosen to select the best model.",
      as_sup("\nb "),
      "'xp' is an acronym for experience"),
    ref_symbols = " "
  ) %>%

  # Table caption
  set_caption(
    caption = "Table 1. Leave-one-out cross-validation table of the five GAMMs relating hunting success to predator experience.",
    style = "Table Caption"
  ) %>%

  # Font options
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 10, part = "all") %>%
  fontsize(size = 12, part = "footer")

tab1
# ===========================================================================
# ===========================================================================

The relationship between hunting success and cumulative experience differed among predators (Figure 1B-D). Only `r toString(scales::percent(percent_tab1[2,]$percentage, accuracy = 0.1))` of the population had an increase in success from the first match to the last in the model where we did not account for the prey's speed and rank (Appendix S1: Figure S1A). In contrast, accounting for both effects resulted in `r toString(scales::percent(percent_tab1[5,]$percentage, accuracy = 0.1))` of the population increasing its success with experience from the first match to the last. The prey speed alone also mediated individual differences in the relationship between success and experience. The standard deviation of the individual slopes component of the model accounting only for prey rank was equal to 9.72 (9.30, 10.15 $95\%\ CI$), while the one for the model accounting for prey rank and speed was equal to 3.36 (3.04, 3.71 $95\%\ CI$), indicating that individual differences in the development of expertise decreased when accounting for prey speed.

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Median posterior predictions of the development of predator hunting expertise. The predators' hunting success (i.e. the probability of consuming the four prey) is on the y axis, and the predators' cumulative experience (i.e. the number of matches played prior to each observation) is on the x axis. Panels A and C show the development of expertise for the average individual with the vertical dashed lines on the left representing the lowest predicted values. For panel A, the right-side vertical dashed line shows the highest predicted success. For panel C, the right-side dashed line represents the point on the curve where success was optimized, which we calculated using the finite differences method to obtain the first derivative of the predicted values. Panels B and D show among individual differences in the development of expertise, with each curve representing an individual predator. (A-B) GAMM where we control for the prey rank (C-D) GAMM where we control for the prey rank and the speed of the prey group."}

knitr::include_graphics(
  here::here("outputs", "04_outputs_figures", "figure1.png")
)

>>>>

---
title: "Individual foraging specialization and success change across experience in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

@@

The capacity of predators to match their tactic to their prey and to optimize their skills at implementing a given tactic are expected to drive the outcome of predator-prey interactions. Hence, successive interactions of predators with their prey may result in increased flexibility in tactic use or in individual foraging specialization. Yet, there are limited empirical assessments showing links between past experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. Here, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how individual predator foraging specialization and success developed across repeated interactions with their prey. We found that 68% of predators became either increasingly specialized by always moving at a fast pace, or flexible by transitioning between slow and fast speeds. The predators' strategies were partially matched to their prey's speed, suggesting that changes in hunting behaviour were driven by repeated encounters with their prey. Flexible and specialist foragers achieved similar success overall. Hence, our findings suggest that experience may promote behavioural diversification in predator-prey systems.

@@

Predators express a diverse array of foraging tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing and fast movement speeds are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions and may have larger implications for community and ecosystem dynamics [@wooster.etal2023].

As predators acquire experience, theory suggests that individuals should gradually specialize in a foraging tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each attempt [@tinker.etal2009;@dukas2019]. This should occur when predators learn to hunt in environments where resources are predictable (or stable) because they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. Such conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning could increase behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting attempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. While there can be costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathot.etal2012]. Nonetheless, complex feedback between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@stephens1993;@ishii.shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences are driven by resource fluctuations [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. For example, @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Prey using fast escapes are also more difficult to hunt [@walker.etal2005;@kelley.magurran2011;@martin.etal2022], and predators that encounter such prey more often should benefit from specializing in fast pursuits. Therefore, if predators hunting in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both achieve similar hunting success and coexist [@phillips.etal2017]. Empiricists should thus investigate the link between foraging specialization and success across time and contexts.

In this study, we investigate how repeated experience shapes predator foraging specialization and success using data from players in *Dead by Daylight*. In this videogame, four prey players forage for resources while avoiding predation by a fifth player. The predator population is composed of individuals that either move at slow speeds and ambush or hunt at high speeds (i.e. mean movement speed describes an ambush-cursorial continuum of hunting tactics) [@@fraserfranco.etal2022]. We also observed that predators acquire expertise through extensive practice, which is driven by the movement of the prey [@fraserfranco.etal2024]. Indeed, the prey can increase their chances of survival by moving faster in the virtual environment [@fraserfranco.etal2022;@fraserfranco.etal2024].

We define the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches. Low IIV describes individual foraging specialization (i.e. always using an ambush or cursorial tactic), whereas high IIV describes a flexible use of foraging tactics (i.e. predators switching between an ambush and cursorial tactic). Here, our objective is to assess how foraging specialization emerges throughout experience at the population and individual level. If all predators encounter similar groups of prey, we predict that individuals (and thus the population) should all specialize on moving at similar speeds. In contrast, if all predators encounter varying groups of prey, then they should converge towards flexible speeds. In both scenarios, differences among individuals in IIV across experience should be low (i.e. similar individual foraging specialization), whereas the population variance would either decrease (specialization) or increase (flexibility). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, we expect predators that encountered similar groups of prey across experience to specialize on similar speeds, while predators that encountered heterogeneous groups across matches should adopt flexible movement speeds, resulting in an increase in among individual differences in IIV with experience. If we detect such prey-dependent fine-tuning with experience, then all hunters along the flexible-specialist continuum should attain equal success.

@@

*DBD* is an asymmetric multiplayer survival game developed by Behaviour Interactive Inc. In this game, participants take on roles as either a predator or one of the prey, with each match consisting of one predator and four prey. A matchmaking system pairs groups of players with similar skill such that individuals should all have similar experience. The predator's primary goal is to hunt and capture the prey, while the prey aim to locate and collect resources in the form of power generators. When all generators are activated, the prey can escape through one of two exit gates and win the match. A match concludes when the predator eliminates all remaining prey that have not escaped or when the final prey successfully exits the virtual environment. Players, whether predator or prey, select avatars with distinct abilities tailored to specific playstyles (e.g. stealthy vs evasive prey, or stalking vs patrolling predators). During the study period, the game featured `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments in the game vary in size and structural complexity, combining fixed and procedurally generated elements such as vegetation, buildings, and maze-like structures. Troughout the study period, there were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for gameplay.

@@

The videogame company provided six months of gameplay data, encompassing all player activity from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We focused exclusively on matches played in "Online" mode, where players were strangers to one another. We excluded matches where players were inactive, as indicated by mean movement speeds that were zero or nearly zero. Additionally, matches where players were suspected of hacking or engaging in unintended gameplay behavior were also removed based on our knowledge of the game mechanics. We then selected a cohort of players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our dataset includes `r toString(format(length(unique(dat$predator_id))))` unique predator players, collectively accounting for `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The number of matches played by predator players ranged from `r toString(format(min(dat$total_xp_pred)))` to `r toString(format(max(dat$total_xp_pred)))`. Match durations varied between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` minutes, with an average duration of `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` minutes. The following details were recorded for each match: the anonymous player ID, the chosen predator avatar, the game environment, the predator's level of experience, their speed, the average speed of the prey group they encountered, and the average rank of those prey (used as a proxy for prey skill). The game's ranking system, designed to match players of similar skill levels (https://deadbydaylight.fandom.com/wiki/Rank), was incorporated into our analysis to detect any relationship between predator experience and foraging success [see @fraserfranco.etal2024].

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we categorized predators for each match based on their cumulated experience [see @fraserfranco.etal2024]. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 500). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

@@

We performed our analyses under a Bayesian framework. We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

### Model specifications

We fitted the model in R (version 4.1.2) using Hamiltonian Monte Carlo (HMC) sampling with the package "brms" version 2.16.3 [@burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.4.0 [@gabry.cesnovar2021] as the back-end for estimation (cmdstan installation version 2.28.2). We ran te model on on Cedar (Operating system: CentOS Linux 7), a computer cluster maintained by the Digital Research Alliance of Canada (https://docs.alliancecan.ca/wiki/Cedar).

We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 used as warm up iterations (yielding 1000 posterior samples for each parameter). We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

After inspection of the results, we found that key parameter values for our hypotheses were in the direction of our predictions, but they were largely driven by the very slow speed of one predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix 2.

@@

@@

# Open percentages table
path <- file.path(here::here("outputs", "03_outputs_model-processing"))

percent_tab2 <- data.table(readRDS(file.path(path, "DHMLM_percentages.rds")))
